Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-07-06 manuscript/manuscript.tex
draft 2 of ms

where $x_{jq,t}$ is the log of species' $j$ proportional cover in
quadrat $q$ at time $t$ and all other parameters are as in the
individual-level growth model (Eq. \#). The log normal likelihood
includes a truncation (T{[}0,1{]}) to ensure that predicted values do
not exceed 100\% cover.

@@

Our quad-based model (QBM) perfectly mirrors its statistical description
(Eq. \#). We use the same approach for drawing parameter values as
described for the IPM.

@@

Equilibrium cover from both models was sensitive to climate (Fig. 3a-d).
The IPM projected percent changes in equilibrium cover from -3 - 8\% for
\emph{B. gracilis}, -4 - 3\% for \emph{H. comata}, -15 - 9\% for
\emph{P. smithii}, and -17 - 53\% for \emph{P. secunda}. The QBM
projected opposite and greater percent changes in equilibrium cover for
\emph{B. gracilis} (-63 - 30\%) and \emph{H. comata} (-50 - -18\%; Fig.
3a-b). For \emph{P. smithii}, the QBM projected opposite changes in
equilibrium cover than the IPM, but of similar magnitude (-5 - 6\%; Fig.
3c). \emph{P. secunda} was the only species that the IPM and QBM made
projections of the same sign and somewhat similar magnitude (Fig. 3d).
As expected based on model validation (Table 1), QBM projections were
more uncertain than IPM projections for all species except \emph{P.
smithiii} (Fig. 3e-h).

The response of a population to climate change is a result of the
aggregate effects of climate on individual vital rates. Since the IPM
approach relies on vital rate regressions, we were able to quantify the
sensitivity of each vital rate in isolation and in pairwise
combinations. Species showed similar trends (Fig. 4). Growth was the
most sensitive vital rate for all species, showing a negative response
to increased precipitation, and stronger positive response to increased
temperature, and a mostly positive response when both climate factors
are increased (Fig. 4). \emph{B. gracilis} survival rates were sensitive
to temperature, resulting in an increase in plant cover under increased
temperature (Fig. 4a). In isolation, recruitment and survival were
insensitive to climate factors for \emph{H. comata} (Fig. 4b). Survival
and recruitment of \emph{P. smithii} were both sensitive, negatively, to
temperature and precipitation (Fig. 4c). \emph{P. secunda} equilibrium
cover was sensitive to the climate effects on survival and recruitment,
showing a negative effect on both vital rates for increased precipition,
but a strong positive effect on survival with increased temperature
(Fig. 4d). The climate impact of recruitment on equilibrium cover was
negative for precipitation and temperature increases (Fig. 4d).

@@

Perhaps the greatest challenge for ecology in the 21st century is to
forecast the impacts of environmental change (Clark et al. 2001). To do
so requires sophisticated modeling approaches that fully account for
uncertainty and variability in the ecological process and associated
parameters. This requires large amounts of data collected over large
spatio-temporal extents. State-of-the-science modeling techniques cannot
overcome data limitations. Such is the case for many population models.

@@

Using individual and aggregated forms of the same dataset, we were able
to directly compare a traditional demographic modeling approach to a
population model based on aggregated data. Our quad-based model (QBM) is
based on percent cover data and so is in the spirit of
density-structured models. In terms of each model's forecasting ability,
the IPM outperformed the QBM (Table 1). This is unsurprising since we
expected to lose some precision at the aggregated level. However, the
underwhelming performance of the QBM could call in the question
forecasts that differ from the IPM.

@@

Given the superior ability of the IPM to predict out of sample
observations (Table 1), we have no choice but to conclude it is the
superior model. Following that logic, we can only assume that, at least
contingent on the data in hand, the IPM is producing the correct
forecasts to climate perturbations. The QBM failed to match IPM
forecasts, implying that detailed demographic data may be necessary to
accurately detect climate signals that are utlimately important at the
population level. This result further confirms related work on the
importance of individual variability on population level responses to
exogenous drivers (CITATIONS).

An important caveat to the above conclusion is that our comparison of
the IPM and the QBM

This result is not as dire as it seems when we consider model
uncerainty. Yes, the QBM failed to match IPM forecasts under perturbed
climate, but the uncertainties around those forecasts were so great
(Fig. 3e-h) that we cannot say that the two model's forecasts were
different with any level of confidence. This offers hope that with
``enough'' population level data one will be able to reduce parameter
uncertainty and thus model forecasts. This remains an open question.

@@

Our intuition was that species with strong climate-dependence on vital
rates not well resolved at the aggregate level would result in different
forecasts from the two models. For example, survival is very size
dependent: smaller individuals have a higher probability of death (Chu
and Adler 2014). At the same time, a single small individual contributes
relatively little to percent cover estimates at the plot scale. So, if
survival of individuals was positively impacted by temperature
increases, for example, we would expect to detect this signal in the
individual level data but not in the aggregate level data.
\textbf{PETER: do you think this is the right direction to go with
this?}

@@

Our goal was not make any explicit forecast for the future state of
these populations based on predicted climate change. But our results
highlight the state of affairs in ecology when it comes to forecasting
the impacts of climate change. The analysis we conducted here could be
considered, with some exceptions of course, at the forefront of
ecological forecasting in terms of the statistical approach employed
(hierarchical Bayesian), the type of population model we used
(stochastic IPM with parameter uncertainty), and the amount of data we
had at our disposal (14 years of individual-level data). Yet, model
predictions proved so uncertain that any forecast, when bounded with
uncertainty, would be at best not useful and at worst meaningless. How
might we improve on this state of affairs?

First, forecasts could be improved by matching the spatial scale of
predictor variables with the spatial scale of observations. One of the
major limitations of the models we fit here is that the climate data are
at a much larger scale than the plot and individual level observations
of plant cover and size. Climate covariates only vary by year, with no
spatial variability within years. Thus, even if we fit models to
individual level data, we are missing the key interaction point between
weather and individual plants. \textbf{PETER: any good citations here
for decent population models fit with fine-scale climate/environmental
data??}

Second, accurately detecting climate signals will take even longer time
series. Recent theoretical work on detecting climate signals in noisy
data suggests that even advanced approaches to parameter fitting like
LASSO and random forest require 20-25 year time series (Teller et al.,
in review).

@@

This work is not a critique of density-structured population models.
Instead, we consider our results a cautionary tale. In some cases and
for certain species, population models based on aggregated data may
prove useful and unbiased. But, as our work shows, that is not a
foregone conclusion when considering the impact of exogenous drivers
like climate on population states. Thus, forecasts from aggregate level
models should be viewed with caution and should never be unaccompianed
by uncertainty. As we have shown, uncertainty around forecasts was large
for both model types. Which leads us to our most pessimistic conclusion:
even with 14 years of detailed demographic data and sophisticated
modeling techniques we failed to produce forecasts with any level of
acceptable uncertainty. In our view, uncertainty of climate change
related forecasts can be reduced by (1) longer time series and (2)
climate covariates that match the scale of inference (e.g., plot rather
than landscape level climate/weather metrics). Do we need detailed
demographic data to forecast the impacts of climate change on
populations? Probably.

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_3.pdf}
\caption{Proportional change in species' mean cover caused by a 1\%
increase in observed precipitation (+ppt), temperature (+temp), or both
(+ppt\&temp) as predicted by the individual-based IPM and the
aggregate-based QBM. Top panels show the mean predicted proportional
change in cover. Lower panels show the sensitivity of equilibrium cover
simulated from the IPM to each climate scenario applied to individual
and combined vital rates. For example, the points associated with G show
the median cover from IPM simulations where a climate perturbation is
applied only to the growth regression climate covariates.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_4.pdf}
\caption{Equilibrium cover and 90\% quantiles around the mean prediction
when model and parameter uncertainty are propogated through the
simulation phase. Climate simulations are as in Figure 3.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_5.pdf}
\caption{Correlations between QBM and IPM estimates of climate effects.
We ignore sizeXclimate interactions since these are not directly
comparable across model types. The QBM does not have multiple vital
rates, so its values are repeated across panels within each species.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_6.pdf}
\caption{Effect of quadrat sample size on the precision (standard
deviation) of main climate effect estimates in the QBM. Increasing the
number of quadrats results in diminishing returns in terms of parameter
certainty. Light dashed lines show individual climate effects at five
quadrat sample sizes. Thick dark lines are inverse gaussian fits showing
the mean effect of increasing quadrat sample size on parameter
precision.}
\end{figure}

@@

@@

>>>>

where $x_{jq,t}$ is the log of species' $j$ proportional cover in
quadrat $q$ at time $t$ and all other parameters are as in the
individual-level growth model (Eq. 3). The log normal likelihood
includes a truncation (T{[}0,1{]}) to ensure that predicted values do
not exceed 100\% cover.

@@

Our quad-based model (QBM) perfectly mirrors its statistical description
(Eq. 8-9). We use the same approach for drawing parameter values as
described for the IPM.

@@

Equilibrium cover from both models was sensitive to climate (Fig. 3a-d).
The IPM projected percent changes in equilibrium cover from -3 - 8\% for
\emph{B. gracilis}, -4 - 3\% for \emph{H. comata}, -15 - 9\% for
\emph{P. smithii}, and -17 - 53\% for \emph{P. secunda}. The QBM
projected opposite and greater percent changes in equilibrium cover for
\emph{B. gracilis} (-63 - 30\%) and \emph{H. comata} (-50 - -18\%; Fig.
3a-b). For \emph{P. smithii}, the QBM projected opposite changes in
equilibrium cover than the IPM, but of similar magnitude (-5 - 6\%; Fig.
3c). \emph{P. secunda} was the only species that the IPM and QBM made
projections of the same sign and somewhat similar magnitude (Fig. 3d).

The response of a population to climate change is a result of the
aggregate effects of climate on individual vital rates. Since the IPM
approach relies on vital rate regressions, we were able to quantify the
sensitivity of each vital rate in isolation and in pairwise
combinations. Species showed similar trends (Fig. 3e-h). Growth was the
most sensitive vital rate for all species, showing a negative response
to increased precipitation, and stronger positive response to increased
temperature, and a mostly positive response when both climate factors
are increased (Fig. 3e-h). \emph{B. gracilis} survival rates were
sensitive to temperature, resulting in an increase in plant cover under
increased temperature (Fig. 3e). In isolation, recruitment and survival
were insensitive to climate factors for \emph{H. comata} (Fig. 3f).
Survival and recruitment of \emph{P. smithii} were both sensitive,
negatively, to temperature and precipitation (Fig. 3g). \emph{P.
secunda} equilibrium cover was sensitive to the climate effects on
survival and recruitment, showing a negative effect on both vital rates
for increased precipition, but a strong positive effect on survival with
increased temperature (Fig. 3h). The climate impact of recruitment on
equilibrium cover was negative for precipitation and temperature
increases (Fig. 3h).

Forecasts based on 1\% climate changes were extremely uncertain when we
considered model error and parameter uncertainty (Fig. 4). As expected
based on model validation (Table 1), QBM projections were more uncertain
than IPM projections for all species except \emph{P. smithiii} (Fig. 4).

@@

Perhaps the greatest challenge for ecology in the 21st century is to
forecast the impacts of environmental change (Clark et al. 2001, Petchey
et al. 2015). To do so requires sophisticated modeling approaches that
fully account for uncertainty and variability in the ecological process
and associated parameters (Luo et al. 2011). This requires large amounts
of data collected over large spatio-temporal extents.
State-of-the-science modeling techniques cannot overcome data
limitations. Such is the case for many population models.

@@

Using individual and aggregated forms of the same dataset, we were able
to directly compare a traditional demographic modeling approach to a
population model based on aggregated data. Our quad-based model (QBM) is
based on percent cover data and so is in the spirit of
density-structured models. In terms of each model's forecasting ability,
the IPM outperformed the QBM (Table 1). This is unsurprising since we
expected to lose some precision at the aggregated level. However, the
underwhelming performance of the QBM could call into the question
forecasts that differ from the IPM.

@@

Given the superior ability of the IPM to predict out of sample
observations (Table 1), we have no choice but to conclude it is the
superior model. Following that logic, we can only assume that, at least
contingent on the data in hand, the IPM is producing the correct
forecasts to climate perturbations. The QBM failed to match IPM
forecasts, implying that detailed demographic data may be necessary to
accurately detect climate signals that are utlimately important at the
population level. This result further confirms related work on the
importance of individual variability on population level responses to
exogenous drivers (Clark et al. 2011a, 2011b, 2012, Galván et al. 2014).

@@

We can think of two reasons why the IPM outperformed the QBM. First, the
quadrat level data has a much reduced sample size compared to the
individual level data. In an ideal world we would have compared the IPM
and QBM using data collected over the same amount of person hours, not
just the same number of quadrats. Then the sample size of the quadrat
level data would be much greater and carry more statistical power. To
address this limitation in our work we fit the QBM statistical model
(Eq. 8-9) with different numbers of quadrats to see the effect of sample
size on the precision of climate effect estimates. It appears that
including additional quadrats leads to rapidly diminishing returns in
terms of parameter precision (Fig. 5). Thus, while sample size surely
plays some role, we do not think it is the main driver of the difference
between the IPM and QBM.

The second reason the IPM could have outperformed the QBM is that the
population level model is in fact missing important climate effects that
act on individual vital rates, rather than population growth. Our
intuition was that species with strong climate-dependence on vital rates
not well resolved at the aggregate level would result in different
forecasts from the two models. For example, survival is very size
dependent: smaller individuals have a higher probability of death (Chu
and Adler 2014). At the same time, a single small individual contributes
relatively little to percent cover estimates at the plot scale. So, if
survival of individuals was positively impacted by temperature
increases, for example, we would expect to detect this signal in the
individual level data but not in the aggregate level data. To see if
this is the case we can regress climate effects from each vital rate
statistical model at the individual level against the same climate
effects from the QBM statistical model (Fig. 6).

In general, the QBM climate effects are most correlated with climate
effects from the growth model at the individual level (Fig. 6). In no
case does the QBM statistical model have strong correlations across all
three vital rates (Fig. 6). Thus, for each species the QBM is
``missing'' climate signals associated with at least one vital rate.
This has large impacts on predictions of long term population dynamics,
as seen in our equilibrium simulations (Fig. 3a-d).

@@

Our goal was not make any explicit forecast for the future state of
these populations based on predicted climate change. But our results
highlight the state of affairs in ecology when it comes to forecasting
the impacts of climate change. The analysis we conducted here could be
considered, with some exceptions of course, at the forefront of
ecological forecasting in terms of the statistical approach employed
(hierarchical Bayesian), the type of population model we used
(stochastic IPM with parameter uncertainty), and the amount of data we
had at our disposal (14 years of individual-level data). Yet, model
predictions proved so uncertain that any forecast, when bounded with
model and parameter uncertainty, would be at best not useful and at
worst meaningless. How might we improve on this state of affairs?

First, forecasts could be improved by matching the spatial scale of
predictor variables with the spatial scale of observations. One of the
major limitations of the models we fit here is that the climate data are
at a much larger scale than the plot and individual level observations
of plant cover and size. Climate covariates only vary by year, with no
spatial variability within years. Thus, even if we fit models to
individual level data, we are missing the key interaction point between
weather and individual plants (Clark et al. 2011b).

Second, accurately detecting climate signals will take even longer time
series. Recent theoretical work on detecting climate signals in noisy
data suggests that even advanced approaches to parameter fitting like
LASSO, functional linear models (splines), and random forest require
20-25 year time series (Teller et al., in review). Alternatively, as we
suggest above, Teller et al. (in review) also find that matching the
scale of the response and predictors improves estimate precision.

Third, ecologists as a community need to get serious about reporting
uncertainty. There is a strong culture around explicitly considering
model uncertainty, but parameter uncertainty is often ignored. In some
cases this is because the easiest statistical methods do no make
propogating parameter uncertainty easy. But even using Bayesian
approaches that allow straightforward integration of model fitting and
forecasting (Hooten and Hobbs 2014, Hobbs and Hooten 2015) is not simple
when using modeling approaches like integral projection models that
separate the model fitting and simulation stages (Rees and Ellner 2009).
However, as we have done here, it is still possible to include parameter
uncertainty by drawing parameter values from MCMC iterations, taking
care to draw all parameters from the same chain and iteration to account
for their correlations. Only by being honest about our forecasts cen we
begin to produce better ones.

@@

This work is not a critique of density-structured population models. In
some cases and for certain species, population models based on
aggregated data may prove useful and unbiased. However, our work here is
the first comparison of population models based on individual and
aggregated data in a climate change context. Our results confirm
theoretical arguments (Clark et al. 2011b) and empirical evidence (Clark
et al. 2011a, 2012) that individual responses are critical to predicting
species' responses to climate change. Thus, forecasts from aggregate
level models should be viewed with caution and should never be
unaccompianed by uncertainty.

Our results also offer a cautionary tale because uncertainty around
forecasts was large for both model types. Which leads us to our most
pessimistic conclusion: even with 14 years of detailed demographic data
and sophisticated modeling techniques we failed to produce forecasts
with any level of acceptable uncertainty. In our view, uncertainty of
climate change related forecasts can be reduced by (1) longer time
series and (2) climate covariates that match the scale of inference
(e.g., plot rather than landscape level climate/weather metrics). Still,
given the poor performance of the quad-based model, it seems there is no
short cut to producing accurate and precise population forecasts. Do we
need detailed demographic data to forecast the impacts of climate change
on populations? Probably.

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_3.pdf}
\caption{Proportional change in species' mean cover caused by a 1\%
increase in observed precipitation (+ppt), temperature (+temp), or both
(+ppt\&temp) as predicted by the individual-based IPM and the
aggregate-based QBM using mean parameter values. Top panels show the
mean predicted proportional change in cover. Lower panels show the
sensitivity of equilibrium cover simulated from the IPM to each climate
scenario applied to individual and combined vital rates. For example,
the points associated with G show the median cover from IPM simulations
where a climate perturbation is applied only to the growth regression
climate covariates. These simulations also use mean parameter values for
clarity.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_4.pdf}
\caption{Equilibrium cover and 90\% quantiles around the mean prediction
when model error and parameter uncertainty are propogated through the
simulation phase. Climate simulations are as in Figure 3.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_5.pdf}
\caption{Effect of quadrat sample size on the precision (standard
deviation) of main climate effect estimates in the QBM. Increasing the
number of quadrats results in diminishing returns in terms of parameter
certainty. Light dashed lines show individual climate effects at five
quadrat sample sizes. Thick dark lines are inverse gaussian fits showing
the mean effect of increasing quadrat sample size on parameter
precision.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_6.pdf}
\caption{Correlations between QBM and IPM estimates of climate effects.
We ignore sizeXclimate interactions since these are not directly
comparable across model types. The QBM does not have multiple vital
rates, so its values are repeated across panels within each species.}
\end{figure}

@@

Clark, J. S., D. M. Bell, M. H. Hersh, and L. Nichols. 2011a. Climate
change vulnerability of forest biodiversity: Climate and competition
tracking of demographic rates. Global Change Biology 17:1834--1849.

Clark, J. S., D. M. Bell, M. H. Hersh, M. C. Kwit, E. Moran, C. Salk, A.
Stine, D. Valle, and K. Zhu. 2011b. Individual-scale variation,
species-scale differences: Inference needed to understand diversity.

@@

Galván, J. D., J. J. Camarero, and E. Gutiérrez. 2014. Seeing the trees
for the forest: Drivers of individual growth responses to climate in
Pinus uncinata mountain forests. Journal of Ecology 102:1244--1257.

Hobbs, N. T., and M. B. Hooten. 2015. Bayesian Models: A Statistical
Primer for Ecologists. Princeton University PressPrinceton.

Hooten, M. B., and N. T. Hobbs. 2014. A guide to Bayesian model
selection for ecologists. Ecological Monographs 85:3--28.

Luo, Y., K. Ogle, C. Tucker, S. Fei, C. Gao, S. LaDeau, J. S. Clark, and
D. S. Schimel. 2011. Ecological forecasting and data assimilation in a
data-rich era. Ecological Applications 21:1429--1442.

Petchey, O. L., M. Pontarp, T. M. Massie, S. Kéfi, A. Ozgul, M.
Weilenmann, G. M. Palamara, F. Altermatt, B. Matthews, J. M. Levine, D.
Z. Childs, B. J. McGill, M. E. Schaepman, B. Schmid, P. Spaak, A. P.
Beckerman, F. Pennekamp, and I. S. Pearse. 2015. The ecological forecast
horizon, and examples of its uses and determinants. Ecology Letters
18:597--611.
